---
title: "Comparing Tornado Data 1950-1989"
author: "Zoe Schroder"
date: '2023-03-28'
output: html_document
editor_options: 
  chunk_output_type: console
---

Projections you may need: 
```{r}
merc <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
US_LCC <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m no_defs"
```

Coordinate Reference Systems you may need:

https://spatialreference.org/ref/esri/usa-contiguous-lambert-conformal-conic/
```{r}
WGS84 <- 4326
US_LCC <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
```

Install the packages for this project: 

```{r}
suppressMessages(library(sf))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(lutz))
suppressMessages(library(xts))
suppressMessages(library(chron))
suppressMessages(library(sp))
suppressMessages(library(ggplot2))
suppressMessages(library(tmap))

#remotes::install_github("ropensci/USAboundaries")
#remotes::install_github("ropensci/USAboundariesData")
suppressMessages(library(USAboundaries))
```

## Historical Tornado Outbreak Data

Load the Historical Outbreak data into this session. 

`HistBigDayCentroids`: The convex hull (boundary) of all touchdown locations for each big day. 
`HistBigDays.sfdfT`: Complete list of any and all convective days with at least 6 tornadoes. A convective day is defined as a 24-hour period between 6 AM and 6 AM. In total, there are 79 historical big days in the data set. 
`HistBigDayTornadoes`: List of all tornadoes that exist on a Big Day with at least 6 tornadoes. In total, 754 tornadoes occur on a big day. 
`HistGroups.sfdfT`: All historical tornadoes grouped into clusters based on space-time distance. The groups range from 1 day to 3 day events.
`HistTorns50_89`: All historical tornadoes between 1950 and 1989 reported in the Grazulis's Significant Tornadoes 1880 - 1989 book. In total, there are 3,108 tornadoes during this time period. 

```{r}
load("HistoricalOutbreaks.RData")
```


## Modern Tornado Outbreak Data

Load the Modern Outbreak data into this session. 

`ModernBigDayCentroids`: The convex hull (boundary) of all touchdown locations for each big day. 
`ModernBigDays.sfdfT`: Complete list of any and all convective days with at least 6 tornadoes. A convective day is defined as a 24-hour period between 6 AM and 6 AM. In total, there are 108 historical big days in the data set. 
`ModernBigDayTornadoes`: List of all tornadoes that exist on a Big Day with at least 6 tornadoes. In total, 1,062 tornadoes occur on a big day. 
`ModernGroups.sfdfT`: All modern tornadoes grouped into clusters based on space-time distance. In total there are 2415 groups. The groups range from 1 day to 3 day events.
`ModernTorns50_89`: All modern tornadoes between 1950 and 1989 reported in the SPC tornado data set. In total, there are 4,622 tornadoes during this time period. 

```{r}
load("ModernOutbreaks.RData")
```

## Data Analysis: 

```{r}
HistBigDays.sfdfT %>%
  arrange(-nT) %>%
  head(10)
```

```{r}
ModernBigDays.sfdfT %>%
  arrange(-nT) %>%
  head(10)
```
There is a 60% match in the top 10 big days between the modern and historical records. The historical record has 4 days which do not fall within the top 10 in the modern record. However, they are found in the top 20 of the modern record. These include: 

May 31, 1985
April 30, 1954
March 29, 1984
April 3, 1956

Get the state data for the lower 48 using the `us_states` function from the **USAboundaries** package. This will allow the state borders to be added to a map using the *tm_shape* function. 
```{r}
sts <- state.name[!state.name %in% c("Alaska", "Hawaii")]
stateBorders <- us_states(states = sts)
```

Plot the data for the historical big day tornado touchdown locations (`HistBigDayTorns`). Add a map of the United States. 
```{r}
HistTornMap <- (tm_shape(stateBorders) +
  tm_borders(col = "gray70", alpha = 1) +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") +       
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("left", "bottom"), 
            inner.margins = c(.1, .1, .1, .1)) +
tm_shape(HistBigDayTornadoes, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_bubbles(size = 0.1, 
             col = "mag_num", 
             breaks = seq(0, 5, by=1),
             labels = c("F1", "F2", "F3", "F4", "F5"), 
             title.col = "Magnitude") +
      tm_layout("Historical Tornadoes",legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2))
```

Plot the data for the modern big day tornado touchdown locations (`ModernBigDayTorns`). Add a map of the United States.

```{r}
ModernTornMap <- (tm_shape(stateBorders) +
  tm_borders(col = "gray70", alpha = 1) +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") +       
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("left", "bottom"), 
            inner.margins = c(.1, .1, .1, .1)) +
tm_shape(ModernBigDayTornadoes, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_bubbles(size = 0.1, 
             col = "mag", 
             breaks = seq(0, 5, by=1),
             labels = c("F1", "F2", "F3", "F4", "F5"), 
             title.col = "Magnitude") +
      tm_layout("Modern Tornadoes", legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2))
```

```{r}
tmap_arrange(HistTornMap, ModernTornMap)
```


From the Historical Data Set:  
April 3, 1974
ID: 
```{r}
Apr3 <- HistBigDays.sfdfT %>%
  filter(cDate == "1974-04-03", groupNumber == 130)

Apr3centroid <- HistBigDayCentroids.df %>%
  filter(cDate == "1974-04-03", groupNumber == 130) %>%
  as_Spatial()

Apr3tornadoes <- HistBigDayTornadoes %>% 
  filter(cDate == "1974-04-03", groupNumber == 130)

Apr3tornadoes <- Apr3tornadoes %>%
  mutate(Hour2 = ifelse(Hour <= 6, Hour + 24, Hour)) %>%
  as_Spatial
```

**Make a map of the APRIL 3 1974 tornado day. Obtain the state and county boundaries from the `USAboundaries` package. **
```{r}
Apr3tornadoes$Hour2 <- cut(as.numeric(Apr3tornadoes$Hour2), breaks=c(6,12,18,24,30))

HistApr31974 <- (tm_shape(stateBorders) +
               tm_text("state_name") +
  tm_borders(col = "gray70", alpha = 1) +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") +       
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("left", "bottom"), 
            inner.margins = c(.15, .15, .15, .15)) +
#tm_shape(counties.sf) +
#  tm_borders(col = "gray40", alpha = .3) +
#  tm_scale_bar(width = 8, size = 8, color.dark = "gray70") +
  #tm_format("World", legend.position = c("right", "top"),
#                   attr.position = c("right", "top"),
#                   legend.frame = FALSE,
                   #title = "May 30th Tornado Group",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
 #                  inner.margins = c(.2, .2, .2, .2)) +
tm_shape(Apr3, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_borders(col = "black", lwd = 3) +
tm_shape(Apr3tornadoes) +
  tm_symbols(size = 4, col = "Hour2", alpha = 0.8, palette = "BuPu", title.col = "Time [CST]", labels = c("6 to 12", "12 to 18", "18 to 24", "0 to 6"), border.alpha = 0) +
    tm_layout(legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2) +
tm_shape(Apr3centroid, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_symbols(size = 1.25, col = "black", shape = 24)  +
  tm_layout(main.title = "Historical Data", title = "April 3, 1974 \n 93 tornadoes", 
              title.position = c("center", "top"), 
              legend.title.size = 1.4,
              legend.position = c("right", "bottom"), 
              legend.stack = "horizontal",
              legend.frame = FALSE, 
              legend.text.size = 1.2, 
              legend.width = -0.2, 
              title.size = 1.5)
)

HistApr31974
```

From the Modern Data Set: 
April 3, 1974
ID: 
```{r}
Apr3 <- ModernBigDays.sfdfT %>%
  filter(cDate == "1974-04-03", groupNumber == 2519)

Apr3centroid <- ModernBigDayCentroids.df %>%
  filter(cDate == "1974-04-03", groupNumber == 2519) %>%
  as_Spatial()

Apr3tornadoes <- ModernBigDayTornadoes %>% 
  filter(cDate == "1974-04-03", groupNumber == 2519)

Apr3tornadoes <- Apr3tornadoes %>%
  mutate(Hour2 = ifelse(Hour <= 6, Hour + 24, Hour)) %>%
  as_Spatial
```

**Make a map of the APRIL 3 1974 tornado day. Obtain the state and county boundaries from the `USAboundaries` package. **
```{r}
Apr3tornadoes$Hour2 <- cut(Apr3tornadoes$Hour2, breaks=c(6,12,18,24,30))

ModApr31974 <- (tm_shape(stateBorders) +
               tm_text("state_name") +
  tm_borders(col = "gray70", alpha = 1) +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") +       
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("left", "bottom"), 
            inner.margins = c(.15, .15, .15, .15)) +
#tm_shape(counties.sf) +
#  tm_borders(col = "gray40", alpha = .3) +
#  tm_scale_bar(width = 8, size = 8, color.dark = "gray70") +
  #tm_format("World", legend.position = c("right", "top"),
#                   attr.position = c("right", "top"),
#                   legend.frame = FALSE,
                   #title = "May 30th Tornado Group",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
 #                  inner.margins = c(.2, .2, .2, .2)) +
tm_shape(Apr3, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_borders(col = "black", lwd = 3) +
tm_shape(Apr3tornadoes) +
  tm_symbols(size = 4, col = "Hour2", alpha = 0.8, palette = "BuPu", title.col = "Time [CST]", labels = c("6 to 12", "12 to 18", "18 to 24", "0 to 6"), border.alpha = 0) +
    tm_layout(legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2) +
tm_shape(Apr3centroid, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_symbols(size = 1.25, col = "black", shape = 24)  +
  tm_layout(main.title = "Modern Data", title = "April 3, 1974 \n 84 tornadoes", 
              title.position = c("center", "top"), 
              legend.title.size = 1.4,
              legend.position = c("right", "bottom"), 
              legend.stack = "horizontal",
              legend.frame = FALSE, 
              legend.text.size = 1.2, 
              legend.width = -0.2, 
              title.size = 1.5)
)

ModApr31974
```

```{r}
tmap_arrange(HistApr31974, ModApr31974)
```

```{r}
 HistBigDayTornadoes %>%
  group_by(mag) %>%
  summarize(test = n())

 ModernBigDayTornadoes %>%
  group_by(mag) %>%
  summarize(test = n())
```
Magnitude & Historical Tornadoes & Modernn Tornadoes \\
F2 & 1199 & 1890 \\
F3 & 550 & 688 \\
F4 & 217 & 228 \\
F5 & 24 & 31 \\

By Decade
```{r}
HistBigDays.sfdfT <- HistBigDays.sfdfT %>% 
  mutate(decade = floor(Year/10)*10) 
ModernBigDays.sfdfT <- ModernBigDays.sfdfT %>% 
  mutate(decade = floor(Year/10)*10) 

tester <- HistBigDays.sfdfT %>%
  group_by(decade) %>%
  summarize(tottorn = sum(nT),
            tornado = tottorn) %>%
  mutate(label = "Historical Tornadoes")

tester2 <- ModernBigDays.sfdfT %>%
  group_by(decade) %>%
  summarize(tottorn = sum(nT)) %>%
  mutate(label = "Modern Tornadoes",
         tornado = tottorn - tester$tornado)

data <- rbind(tester,tester2)

```

```{r}
ggplot(data, aes(x = decade, y = tornado, fill = forcats::fct_rev(label))) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  xlab("Decade") +
  ylab("Number of Tornadoes") +
  scale_fill_manual(values=c( "red", "blue")) +
  theme(legend.position="bottom", legend.title=element_blank(), axis.text = element_text(size = 14), axis.title = element_text(size = 16), legend.text = element_text(size = 16)) +
  scale_y_continuous(expand = c(0, 0), name="Number of Tornadoes", limits=c(0, 900), breaks = seq(0, 900, 100))
```

