---
title: "Comparing Tornado Data 1950-1989"
author: "Zoe Schroder"
date: '2023-03-28'
output: html_document
---

Projections you may need: 
```{r}
merc <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
US_LCC <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m no_defs"
```

Coordinate Reference Systems you may need:
```{r}
WGS84 <- 4326
```

Install the packages for this project: 

```{r}
suppressMessages(library(sf))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(lutz))
suppressMessages(library(xts))
suppressMessages(library(chron))
```

## Historical Tornado Data

Load the original Significant Tornado data into this session. 
```{r, eval = FALSE}
load("OriginalHistoricalTornadoData.RData")
```

Save the line (tornado track) geometry for later use. 
```{r, eval = FALSE}
SigTor_geom <- SigTor$geometry 
```

Calculate the casualties, path of the damage area, and abbreviation of the month. 
Separate the date: 
```{r, eval = FALSE}
SigTor <- SigTor %>%
  mutate(DateTime = as.POSIXct(paste(date, time), format="%Y-%m-%d %H:%M:%S"), 
         cas = inj + fat,
         AreaPath = len * wid,
         Ma = factor(month.abb[mo], levels = month.abb[1:12])) %>%
  sf::st_sf()
dim(SigTor)[1]
```

Compute the Energy Dissipation (ED) where $A_p$ is the area of the path, $\rho$ is area density [1 kg/m^3]  $v_j$ is the midpoint wind speed for each rating, and $w_j$ is the corresponding fraction of path area by EF rating. With no upper bound on the EF5 wind speeds, the midpoint wind speed is set at 97 m~s$^{-1}$ (7.5 m~s$^{-1}$ above the threshold wind speed consistent with the EF4 midpoint speed relative to its threshold)
```{r, eval = FALSE}
perc <- c(1, 0, 0, 0, 0, 0, 
          .772, .228, 0, 0, 0, 0,
          .616, .268, .115, 0, 0, 0,
          .529, .271, .133, .067, 0, 0,
          .543, .238, .131, .056, .032, 0,
          .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
mag_num <- as.numeric(substr(SigTor$mag,2,2))
ef <- mag_num + 1 
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
SigTor <- SigTor %>%
  mutate(ED = EW3 * AreaPath) #Units of ______
``` 

Get Tornado touchdown (point) data: 
```{r, eval = FALSE}
SigTor.sfdf <- st_drop_geometry(SigTor)

SigTor.sfdf <- SigTor.sfdf %>%
  st_as_sf(coords = c("slon", "slat"), crs = 4326)
slat <- SigTor$slat
slon = SigTor$slon

SigTor.sfdf <- cbind(SigTor.sfdf, slat, slon)
```

Calculate the Convective day!:
```{r, eval = FALSE}
SigTor.sfdf <- SigTor.sfdf %>%
  mutate(cDateTime = DateTime - as.difftime(6, unit = "hours"),
         hour = hour(DateTime),
         cDate = as.Date(as_datetime(ifelse(hour < 6, (DateTime - 86400), cDateTime), tz = Sys.timezone()))) %>%
  sf::st_sf()
```

Calculate the NARRtime, NARRday, and NARRZtime for each report. 
```{r, eval = FALSE}
SigTor.sfdf$Zulu_DateTime <- force_tz(SigTor.sfdf$DateTime, tzone = "UTC")
NARRtime <- align.time(SigTor.sfdf$Zulu_DateTime, n = (60 * 60 * 3)) + 3600 *3 ###Had to remove the -3600*3
NARRday = format(as.POSIXct(strptime(NARRtime,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%Y/%m/%d")
NARRZtime = format(as.POSIXct(strptime(NARRtime,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%H")

SigTor.sfdf <- cbind(SigTor.sfdf, NARRday, NARRZtime)
```

Establish the geometry for a tornado tracks dataset using the original geometry from the Historical Tornadoes data. 
```{r, eval = FALSE}
#Tornado Tracks
SigTornTracks<- st_as_sf(cbind(st_drop_geometry(SigTor.sfdf), SigTor_geom))
SigTorns <- SigTor.sfdf
```


## Modern Tornado Data

Load the Modern Tornado Record in to the R workspace. The modern tornado data comes from: https://www.spc.noaa.gov/wcm/#data. This data is from 1950 to 2021
```{r, eval = FALSE}
Torn50_21.spdf <- read.csv(file = "1950-2021_actual_tornadoes.csv")
```

Set up the data for the  Modern Tornado Record (as points): 
```{r, eval = FALSE}
ModernTorns <- Torn50_21.spdf
sp::coordinates(ModernTorns) <- ~ slon + slat
ModernTorns <- st_as_sf(ModernTorns)
st_crs(ModernTorns) <- WGS84
```

Set up the data for the  Modern Tornado Record (as tracks): 

```{r, eval = FALSE}
ModernTornTracks <- Torn50_21.spdf
sp::coordinates(ModernTornTracks) <- ~ slon + elon + slat  + elat
ModernTornTracks <- st_as_sf(ModernTornTracks)
st_crs(ModernTornTracks) <- WGS84
```

Re-add the start lat and start lon to the Modern Tracks data
```{r, eval = FALSE}
slon <- ModernTornTracks$slon
slat <- ModernTornTracks$slat
ModernTorns <- cbind(ModernTorns, slon, slat)
```

```{r, eval = FALSE}
#save(ModernTorns, ModernTornTracks, file = "ModernTornadoes.RData")
```

## Load the Historical Tornado Data: 

This data set includes the **HistTorns** and the **HistTornTracks** data. The **HistTorns** is the point data for the tornado genesis location. The **HistTornTracks** is the line data for the entire track (path) of the tornado.

```{r}
load("HistoricalTornadoes.RData")
```

## Load the Modern Tornado Data: 

This data set includes the **ModernTorns** and the **ModernTornTracks** data for all recorded tornadoes between 1950 and 2021. The **ModernTorns** is the point data for the tornado genesis location. The **ModernTornTracks** is the line data for the entire track (path) of the tornado.

```{r}
load("ModernTornadoes.RData")
```


## Data Cleaning: 

Remove tornadoes in Hawaii, Alaska, and Puerto Rico and those occurring before 1994. That year marks the beginning of comprehensive WSR-88D radar. For missing EF ratings use the modification rules (if/else) defined here: https://www.spc.noaa.gov/wcm/OneTor_F-scale-modifications.pdf 
```{r}
All_Tornadoes <- Tor.sfdf %>%
  filter(yr >= 1994,
         !st %in% c("AK", "PR", "HI")) %>%
  mutate(mag = ifelse(mag == -9 & len <= 5, 0, mag),
         mag = ifelse(mag == -9 & len > 5, 1, mag))
```

## Define clusters for Historical Tornadoes:



## Define clusters for Modern Tornadoes:

