---
title: "Comparing Tornado Data 1950-1989"
author: "Zoe Schroder"
date: '2023-03-28'
output: html_document
editor_options: 
  chunk_output_type: console
---

For publication in **Theoretical and Applied Climatology**

Projections you may need: 
```{r}
merc <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
US_LCC <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m no_defs"
```

Coordinate Reference Systems you may need:

https://spatialreference.org/ref/esri/usa-contiguous-lambert-conformal-conic/
```{r}
WGS84 <- 4326
US_LCC <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
```

Install the packages for this project: 

```{r}
suppressMessages(library(sf))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(lutz))
suppressMessages(library(xts))
suppressMessages(library(chron))
suppressMessages(library(sp))
suppressMessages(library(ggplot2))
suppressMessages(library(tmap))

#remotes::install_github("ropensci/USAboundaries")
#remotes::install_github("ropensci/USAboundariesData")
suppressMessages(library(USAboundaries))
suppressMessages(library(USAboundariesData))

suppressMessages(library(ggspatial))
suppressMessages(library(sfhotspot))
suppressMessages(library(tidyverse))
suppressMessages(library(gridExtra))
```

## Historical Tornado Outbreak Data

Load the Historical Outbreak data into this session. 

`HistBigDayCentroids`: The convex hull (boundary) of all touchdown locations for each big day. 
`HistBigDays.sfdfT`: Complete list of any and all convective days with at least 6 tornadoes. A convective day is defined as a 24-hour period between 6 AM and 6 AM. In total, there are 79 historical big days in the data set. 
`HistBigDayTornadoes`: List of all tornadoes that exist on a Big Day with at least 6 tornadoes. In total, 754 tornadoes occur on a big day. 
`HistGroups.sfdfT`: All historical tornadoes grouped into clusters based on space-time distance. The groups range from 1 day to 3 day events.
`HistTorns50_89`: All historical tornadoes between 1950 and 1989 reported in the Grazulis's Significant Tornadoes 1880 - 1989 book. In total, there are 3,108 tornadoes during this time period. 

```{r}
load("HistoricalOutbreaks.RData")
```


## Modern Tornado Outbreak Data

Load the Modern Outbreak data into this session. 

`ModernBigDayCentroids`: The convex hull (boundary) of all touchdown locations for each big day. 
`ModernBigDays.sfdfT`: Complete list of any and all convective days with at least 6 tornadoes. A convective day is defined as a 24-hour period between 6 AM and 6 AM. In total, there are 108 historical big days in the data set. 
`ModernBigDayTornadoes`: List of all tornadoes that exist on a Big Day with at least 6 tornadoes. In total, 1,062 tornadoes occur on a big day. 
`ModernGroups.sfdfT`: All modern tornadoes grouped into clusters based on space-time distance. In total there are 2415 groups. The groups range from 1 day to 3 day events.
`ModernTorns50_89`: All modern tornadoes between 1950 and 1989 reported in the SPC tornado data set. In total, there are 4,622 tornadoes during this time period. 

```{r}
load("ModernOutbreaks.RData")
```

## Data Analysis: Tornadoes

Get the state data for the lower 48 using the `us_states` function from the **USAboundaries** package. This will allow the state borders to be added to a map using the *tm_shape* function. 
```{r}
sts <- state.name[!state.name %in% c("Alaska", "Hawaii")]
stateBorders <- us_states(states = sts)
```

Plot the data for the historical big day tornado touchdown locations (`HistBigDayTorns`). Add a map of the United States. 
```{r}
HistTornadoes <- HistTorns50_89 %>%
  filter(mag != "FT", 
         mag !="FN")

HistTornMap <- (tm_shape(stateBorders) +
  tm_borders(col = "gray70", alpha = 1) +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") +       
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("left", "bottom"), 
            inner.margins = c(.05, .05, .1, .05)) + # (S, W, N, E)
tm_shape(HistTornadoes, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_bubbles(size = 0.1, 
             col = "mag_num", 
             breaks = seq(0, 5, by=1),
             labels = c("F1", "F2", "F3", "F4", "F5"), 
             title.col = "Magnitude") +
      tm_layout("SigTor Dataset (Total: 5,692)", legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2))
```

Plot the data for the modern big day tornado touchdown locations (`ModernBigDayTorns`). Add a map of the United States.

```{r}
ModTornadoes <- ModernTorns50_89

ModernTornMap <- (tm_shape(stateBorders) +
  tm_borders(col = "gray70", alpha = 1) +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") +       
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("left", "bottom"), 
            inner.margins = c(.05, .05, .1, .05)) + # (S, W, N, E)
tm_shape(ModTornadoes, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_bubbles(size = 0.1, 
             col = "mag", 
             breaks = seq(0, 5, by=1),
             labels = c("F1", "F2", "F3", "F4", "F5"), 
             title.col = "Magnitude") +
      tm_layout("ONETOR Dataset (Total: 8,367)", legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2))
```

```{r}
tmap_arrange(HistTornMap, ModernTornMap)
```
`Fig_HistModTorns`

```{r}
 HistTornadoes %>%
  group_by(mag) %>%
  summarize(test = n())

 ModTornadoes %>%
  group_by(mag) %>%
  summarize(test = n())
```
`Tab_TornMag`
Magnitude & Historical Tornadoes & Modern Tornadoes \\
F2 & 4287 & 6306 \\
F3 & 993 & 1636 \\
F4 & 375 & 385 \\
F5 & 33 & 40 \\

By Decade
```{r}
HistTornadoes <- HistTornadoes %>% 
  mutate(decade = floor(Year/10)*10) 
ModTornadoes <- ModTornadoes %>% 
  mutate(decade = floor(Year/10)*10) 

tester <- HistTornadoes %>%
  group_by(decade) %>%
  summarize(tottorn = n(),
            tornado = tottorn) %>%
  mutate(label = "SigTor")

tester2 <- ModTornadoes %>%
  group_by(decade) %>%
  summarize(tottorn = n()) %>%
  mutate(label = "ONETOR",
         tornado = tottorn - tester$tornado)

data <- rbind(tester,tester2)
```

```{r}
ggplot(data, aes(x = decade, y = tornado, fill = label)) + #forcats::fct_rev(label)
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  xlab("Decade") +
  ylab("Number of Tornadoes") +
  scale_fill_manual(values=c( "red", "blue")) +
  theme(legend.position="bottom", legend.title=element_blank(), axis.text = element_text(size = 14), axis.title = element_text(size = 16), legend.text = element_text(size = 16)) +
  scale_y_continuous(expand = c(0, 0), name="Number of Tornadoes", limits=c(0, 2600), breaks = seq(0, 2600, 250))
```
`Fig_TornCounts`

```{r}
histbyyear <- HistTornadoes  %>%
  group_by(yr) %>%
  summarize(hisT = n())

modbyyear <- ModTornadoes %>%
  group_by(yr) %>%
  summarize(modT = n())

test <- merge(as.data.frame(histbyyear), as.data.frame(modbyyear), by = "yr")

missingtorns <- test %>%
  mutate(missingtorn = modT-hisT)
```

```{r}
ggplot(missingtorns, aes(x = yr, y = missingtorn)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(legend.position="bottom", legend.title=element_blank(), axis.text = element_text(size = 14), axis.title = element_text(size = 16), legend.text = element_text(size = 16)) +
  scale_y_continuous(expand = c(0, 0), name="Number of Tornadoes", limits=c(-25, 180), breaks = seq(-25, 175, 25)) +
  scale_x_continuous(expand = c(0, 0), name="Year", limits=c(1949, 1991), breaks = seq(1950, 1991, 5))
```
`Fig_MissingTorns`

KDE of Historical Tornadoes
```{r, eval = FALSE}
torn_density_clip <- HistTornadoes |>
  hotspot_kde(cell_size = 25000, 
              bandwidth_adjust = 0.5) |>
  st_transform(WGS84) |>
  st_intersection(stateBorders)
```

```{r, eval = FALSE}
histkde <- (ggplot() + 
  #annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(fill = kde), 
          data = torn_density_clip, 
          alpha = 0.75,
          colour = NA) +
  geom_sf(data = stateBorders, color = "gray70", fill = NA) +
  geom_sf_label(aes(label = str_wrap(stusps, width = 10)), 
                data = stateBorders, 
                alpha = 0,
                colour = "gray20", 
                lineheight = 1, 
                size = 2.5, 
                label.size = NA) +
  annotation_scale(style = "ticks", location = "bl") +
  labs(title = "Tornadoes in the United States",
        subtitle = "SigTor Data: 1950 - 1989", 
       fill = "Kernel Density \nof Tornadoes") +
  scale_fill_distiller(palette = "PuBu", 
                       direction = 1,
                       breaks = range(pull(torn_density_clip, "kde")),
                       labels = c("lower", "higher")) + 
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(size = rel(0.8), hjust = 0.5)))
```

KDE of Modern Tornadoes
```{r, eval = FALSE}
torn_density_clip <- ModTornadoes |>
  hotspot_kde(cell_size = 25000, 
              bandwidth_adjust = 0.5) |>
    st_transform(WGS84) |>
  st_intersection(stateBorders)
```

```{r, eval = FALSE}
modkde <- (ggplot() + 
  #annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(fill = kde), 
          data = torn_density_clip, 
          alpha = 0.75,
          colour = NA) +
  geom_sf(data = stateBorders, color = "gray70", fill = NA) +
  geom_sf_label(aes(label = str_wrap(stusps, width = 10)), 
                data = stateBorders, 
                alpha = 0,
                colour = "gray20", 
                lineheight = 1, 
                size = 2.5, 
                label.size = NA) +
  annotation_scale(style = "ticks", location = "bl") +
  labs(title = "Tornadoes in the United States",
        subtitle = "ONETOR Data: 1950 - 1989", 
       fill = "Kernel Density \nof Tornadoes") +
  scale_fill_distiller(palette = "PuBu", 
                       direction = 1,
                       breaks = range(pull(torn_density_clip, "kde")),
                       labels = c("lower", "higher")) + 
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(size = rel(0.8), hjust = 0.5)))
```

```{r, eval = FALSE}
grid.arrange(histkde, modkde)
```
`Fig_TornKDE`

```{r}
HistTorn_Mo <- HistTornadoes %>%
  group_by(mo) %>%
  summarize(numTorn = n())

ModTorn_Mo <- ModTornadoes %>%
  group_by(mo) %>%
  summarize(numTorn = n())
```

`Make a figure of the total clusters by month`
```{r}
p0 <- ggplot(HistTorn_Mo, aes(x = as.factor(mo), y = numTorn)) +
  geom_bar(stat = "identity", fill = "gray70") +
  scale_x_discrete(breaks = seq(1, 12, 1), labels = month.abb) +
  scale_y_continuous(limits = c(0, 1800)) +
  coord_polar(start = 0) +
  labs(x = "Month", y = "Number of Tornadoes") +
  theme_minimal() +
  ggtitle("A")

p1 <- ggplot(ModTorn_Mo, aes(x = as.factor(mo), y = numTorn)) +
  geom_bar(stat = "identity", fill = "gray70") +
  scale_x_discrete(breaks = seq(1, 12, 1), labels = month.abb) +
  scale_y_continuous(limits = c(0, 1800)) +
  coord_polar(start = 0) +
  labs(x = "Month", y = "Number of Tornadoes") +
  theme_minimal() +
  ggtitle("B")
```

```{r}
grid.arrange(p0, p1, nrow = 1)
```
`Fig_TornbyMo`

## Data Analysis: Outbreaks


```{r}
HistBigDays.sfdfT %>%
  select(cDate, nT, GroupDayCas) %>%
  arrange(-nT) %>%
  head(10)
```

```{r}
ModernBigDays.sfdfT %>%
  select(cDate, nT, GroupDayCas) %>%
  arrange(-nT) %>%
  head(10)
```
`Tab_Top10Outbreaks`

There is a 60% match in the top 10 big days between the modern and historical records. The historical record has 4 days which do not fall within the top 10 in the modern record. However, they are found in the top 20 of the modern record. These include: 

May 31, 1985
April 30, 1954
March 28, 1984
April 3, 1956

```{r}
Hist_test <- HistBigDays.sfdfT %>% 
  mutate(category = floor(nT/10)*10) 
Mod_test <- ModernBigDays.sfdfT %>% 
  mutate(category = floor(nT/10)*10) 

Hist_test %>%
  group_by(category) %>%
  summarize(numClus = n(), 
            totCas = sum(GroupDayCas))

Mod_test %>%
  group_by(category) %>%
  summarize(numClus = n(), 
            totCas = sum(GroupDayCas))
```
nT    & NumClus & totCas & NumClus & totCas
6-9   &   124   &  9698  &   180   &  11142
10-19 &    57   & 15035  &    89   &  10500
20-29 &     5   &  3261  &    13   &   7270
30-39 &     1   &  3698  &     1   &   3632
40+   &     1   &  5694  &     1   &   5660

From the Historical Data Set:  
April 3, 1974
ID: 
```{r}
Apr3 <- HistBigDays.sfdfT %>%
  filter(cDate == "1974-04-03", groupNumber == 130)

Apr3centroid <- HistBigDayCentroids.df %>%
  filter(cDate == "1974-04-03", groupNumber == 130) %>%
  as_Spatial()

Apr3tornadoes <- HistBigDayTornadoes %>% 
  filter(cDate == "1974-04-03", groupNumber == 130)

Apr3tornadoes <- Apr3tornadoes %>%
  mutate(Hour2 = ifelse(Hour <= 6, Hour + 24, Hour)) %>%
  as_Spatial
```

**Make a map of the APRIL 3 1974 tornado day. Obtain the state and county boundaries from the `USAboundaries` package. **
```{r}
Apr3tornadoes$Hour2 <- cut(as.numeric(Apr3tornadoes$Hour2), breaks=c(6,12,18,24,30))

HistApr31974 <- (tm_shape(stateBorders) +
               tm_text("state_name") +
  tm_borders(col = "gray70", alpha = 1) +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") +       
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("left", "bottom"), 
            inner.margins = c(.15, .15, .15, .15)) +
#tm_shape(counties.sf) +
#  tm_borders(col = "gray40", alpha = .3) +
#  tm_scale_bar(width = 8, size = 8, color.dark = "gray70") +
  #tm_format("World", legend.position = c("right", "top"),
#                   attr.position = c("right", "top"),
#                   legend.frame = FALSE,
                   #title = "May 30th Tornado Group",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
 #                  inner.margins = c(.2, .2, .2, .2)) +
tm_shape(Apr3, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_borders(col = "black", lwd = 3) +
tm_shape(Apr3tornadoes) +
  tm_symbols(size = 4, col = "Hour2", alpha = 0.8, palette = "BuPu", title.col = "Time [CST]", labels = c("6 to 12", "12 to 18", "18 to 24", "0 to 6"), border.alpha = 0) +
    tm_layout(legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2) +
tm_shape(Apr3centroid, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_symbols(size = 1.25, col = "black", shape = 24)  +
  tm_layout(main.title = "SigTor Dataset", title = "April 3, 1974 \n 93 tornadoes", 
              title.position = c("center", "top"), 
              legend.title.size = 1.4,
              legend.position = c("right", "bottom"), 
              legend.stack = "horizontal",
              legend.frame = FALSE, 
              legend.text.size = 1.2, 
              legend.width = -0.3, 
              title.size = 1.5)
)

HistApr31974
```

From the Modern Data Set: 
April 3, 1974
ID: 
```{r}
Apr3 <- ModernBigDays.sfdfT %>%
  filter(cDate == "1974-04-03", groupNumber == 2519)

Apr3centroid <- ModernBigDayCentroids.df %>%
  filter(cDate == "1974-04-03", groupNumber == 2519) %>%
  as_Spatial()

Apr3tornadoes <- ModernBigDayTornadoes %>% 
  filter(cDate == "1974-04-03", groupNumber == 2519)

Apr3tornadoes <- Apr3tornadoes %>%
  mutate(Hour2 = ifelse(Hour <= 6, Hour + 24, Hour)) %>%
  as_Spatial
```

**Make a map of the APRIL 3 1974 tornado day. Obtain the state and county boundaries from the `USAboundaries` package. **
```{r}
Apr3tornadoes$Hour2 <- cut(Apr3tornadoes$Hour2, breaks=c(6,12,18,24,30))

ModApr31974 <- (tm_shape(stateBorders) +
               tm_text("state_name") +
  tm_borders(col = "gray70", alpha = 1) +
  tm_compass(size = 3, fontsize = 1, lwd = 2, color.dark = "gray70") +       
  tm_scale_bar(width = .3, size = 0.8, lwd = 1.75, color.dark = "gray70") +
  tm_layout(legend.bg.color = "white", 
            legend.text.size = .75, 
            attr.position = c("left", "bottom"), 
            inner.margins = c(.15, .15, .15, .15)) +
#tm_shape(counties.sf) +
#  tm_borders(col = "gray40", alpha = .3) +
#  tm_scale_bar(width = 8, size = 8, color.dark = "gray70") +
  #tm_format("World", legend.position = c("right", "top"),
#                   attr.position = c("right", "top"),
#                   legend.frame = FALSE,
                   #title = "May 30th Tornado Group",
                   #title.size = 1.3,
                   #title.position = c("left", "TOP"),
 #                  inner.margins = c(.2, .2, .2, .2)) +
tm_shape(Apr3, is.master = TRUE, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_borders(col = "black", lwd = 3) +
tm_shape(Apr3tornadoes) +
  tm_symbols(size = 4, col = "Hour2", alpha = 0.8, palette = "BuPu", title.col = "Time [CST]", labels = c("6 to 12", "12 to 18", "18 to 24", "0 to 6"), border.alpha = 0) +
    tm_layout(legend.title.size = 1.1,
            legend.position = c("right", "bottom"), 
            legend.stack = "horizontal",
            legend.frame = FALSE, 
            legend.text.size = 1, legend.width = -0.2) +
tm_shape(Apr3centroid, projection = "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") +
  tm_symbols(size = 1.25, col = "black", shape = 24)  +
  tm_layout(main.title = "ONETOR Dataset", title = "April 3, 1974 \n 96 tornadoes", 
              title.position = c("center", "top"), 
              legend.title.size = 1.4,
              legend.position = c("right", "bottom"), 
              legend.stack = "horizontal",
              legend.frame = FALSE, 
              legend.text.size = 1.2, 
              legend.width = -0.3, 
              title.size = 1.5)
)

ModApr31974
```

```{r}
tmap_arrange(HistApr31974, ModApr31974)
```
`Fig_April31974`

```{r}
HistClus_Mo <- HistBigDays.sfdfT %>%
  group_by(Mo) %>%
  summarize(numClus = n())

ModClus_Mo <- ModernBigDays.sfdfT %>%
  group_by(Mo) %>%
  summarize(numClus = n())
```

`Make a figure of the total clusters by month`
```{r}
p0 <- ggplot(HistClus_Mo, aes(x = as.factor(Mo), y = numClus)) +
  geom_bar(stat = "identity", fill = "gray70") +
  scale_x_discrete(breaks = seq(1, 12, 1), labels = month.abb) +
  scale_y_continuous(limits = c(0, 80)) +
  coord_polar(start = 0) +
  labs(x = "Month", y = "Number of Clusters") +
  theme_minimal() +
  ggtitle("A")

p1 <- ggplot(ModClus_Mo, aes(x = as.factor(Mo), y = numClus)) +
  geom_bar(stat = "identity", fill = "gray70") +
  scale_x_discrete(breaks = seq(1, 12, 1), labels = month.abb) +
  scale_y_continuous(limits = c(0, 80)) +
  coord_polar(start = 0) +
  labs(x = "Month", y = "Number of Clusters") +
  theme_minimal() +
  ggtitle("B")
```

```{r}
grid.arrange(p0, p1, nrow = 1)
```
`Fig_ClusbyMo`

## Paired T-Test

Use modern as truth

Set up data frame for paired t tests:
```{r}
SigTor.dat <- HistBigDays.sfdfT %>%
  group_by(Year) %>%
  summarize(bigdays = n(),
            numTorn = sum(nT), 
            avgnT = round(mean(nT)))

Year <- 1951
bigdays = 0
numTorn = 0
avgnT = 0
test <- cbind(Year, bigdays, numTorn, avgnT)

SigTor.dat <- st_drop_geometry(SigTor.dat)
SigTor.dat <- rbind(SigTor.dat, test)

ONETOR.dat <- ModernBigDays.sfdfT %>%
  group_by(Year) %>%
  summarize(bigdays = n(),
            numTorn = sum(nT), 
            avgnT = round(mean(nT)))

ONETOR.dat <- st_drop_geometry(ONETOR.dat)

SigTor.dat <- as.data.frame(SigTor.dat)
ONETOR.dat <- as.data.frame(ONETOR.dat)

SigTor.dat$group <- rep("SigTor",40)
ONETOR.dat$group <- rep("ONETOR",40)

dat2 <- rbind(SigTor.dat, ONETOR.dat)
```

`Null Hypothesis:` There is no relationship between the data. 
`Alternative Hypothesis:` There is a relationship between the data!
```{r}
bartlett.test(bigdays ~ group, data = dat2)
#Test the number of big days each year between HIST and MOD
t.test(bigdays ~ group, data = dat2, paired=TRUE, var.equal = FALSE)
# pvalue: 0.006301
#Reject the null: The means of the two datasets is NOT significantly different!
```

```{r}
#Test the number of tornadoes each year between HIST and MOD
bartlett.test(numTorn ~ group, data = dat2)
t.test(numTorn ~ group, data = dat2, paired=TRUE, var.equal = FALSE)
# pvalue: 0.006587
#Reject the null: The means of the two datasets is NOT significantly different!
```

```{r}
#Test the average number of tornadoes per outbreak each year between HIST and MOD
bartlett.test(avgnT ~ group, data = dat2)
t.test(avgnT ~ group, data = dat2, paired=TRUE, var.equal = FALSE)
# pvalue: 1
```


