---
title: "Comparing Tornado Data 1950-1989"
author: "Zoe Schroder"
date: '2023-03-28'
output: html_document
---

Projections you may need: 
```{r}
merc <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
US_LCC <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m no_defs"
```

Coordinate Reference Systems you may need:

https://spatialreference.org/ref/esri/usa-contiguous-lambert-conformal-conic/
```{r}
WGS84 <- 4326
US_LCC <- "+proj=lcc +lat_1=33 +lat_2=45 +lat_0=39 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
```

Install the packages for this project: 

```{r}
suppressMessages(library(sf))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(lutz))
suppressMessages(library(xts))
suppressMessages(library(chron))
suppressMessages(library(sp))
```

## Historical Tornado Outbreak Data

Load the Historical Outbreak data into this session. 

`HistBigDayCentroids`: The convex hull (boundary) of all touchdown locations for each big day. 
`HistBigDays.sfdfT`: Complete list of any and all convective days with at least 6 tornadoes. A convective day is defined as a 24-hour period between 6 AM and 6 AM. In total, there are 188 historical big days in the data set. 
`HistBigDayTornadoes`: List of all tornadoes that exist on a Big Day with at least 6 tornadoes. In total, 1881 tornadoes occur on a big day. 
`HistGroups.sfdfT`: All historical tornadoes grouped into clusters based on space-time distance. The groups range from 1 day to 3 day events.
`HistTorns50_89`: All historical tornadoes between 1950 and 1989 reported in the Grazulis's Significant Tornadoes 1880 - 1989 book. In total, there are 5,692 tornadoes during this time period. 

```{r}
load("HistoricalOutbreaks.RData")
```


## Modern Tornado Outbreak Data

Load the Modern Outbreak data into this session. 

`ModernBigDayCentroids`: The convex hull (boundary) of all touchdown locations for each big day. 
`ModernBigDays.sfdfT`: Complete list of any and all convective days with at least 6 tornadoes. A convective day is defined as a 24-hour period between 6 AM and 6 AM. In total, there are 759 historical big days in the data set. 
`ModernBigDayTornadoes`: List of all tornadoes that exist on a Big Day with at least 6 tornadoes. In total, 8531 tornadoes occur on a big day. 
`ModernGroups.sfdfT`: All modern tornadoes grouped into clusters based on space-time distance. In total there are 7395 groups. The groups range from 1 day to 3 day events.
`ModernTorns50_89`: All modern tornadoes between 1950 and 1989 reported in the SPC tornado data set. In total, there are 19736 tornadoes during this time period. 

```{r}
load("ModernOutbreaks.RData")
```



